Are populations like a circuit? The relationship between isolation by distance and isolation by resistance

### 2018-08-30 isolation_by_coalescence.tex

The idea of ``resistance distance'' is an important tool in the lanscape genomics toolbox.
Introduced by \citet{mcrae},
it makes use of a mathematical equivalence between random walks and electrical resistance
\citep{resistance}:
the expected time for a continuous-time Markov chain that starts at node $x$
to first hit node $y$ and then return to $x$ (the ``commute time'')
is equal to the effective resistance between $x$ and $y$ in an electrical network
whose conductances are given by the movement rates of the Markov chain \citep{levin}.
Both electrical resistance and commute time between two nodes
average over all possible paths through the network between the two.
For this reason, it is thought that if genetic distances between samples
from different parts of the landscape
are significantly correlated with the effective resistances between those locations,
with local conductances assigned based on the values of some landscape variable,
then that landscape variable is likely a good indicator of the regions of the map 
where gene flow occurs.
For instance,
if the conductance across a grid cell in a discretization of the landscape
is higher in flatter areas,
and this produces effective resistance values 
that are significantly correlated with genetic distances,
then one might conclude that the species in question tends to disperse more readily through flatter areas,
perhaps along river valleys.
\plr{reference some examples}
Other methods \citep{eems} seek to build \emph{de novo} 
a map of conductance values that produces resistances 
most strongly correlated with observed genetic distance,
and then interpret regions of low conductance in the resulting map as barriers to gene flow.

Genetic distances between two genomes sampled from the landscape
do derive from an average across a large number of paths between those points --
the lineages along which each segment of genetic material has been inherited
from their most recent common ancestor.
The more recent this ancestor is,
the smaller the genetic distance will be.
Thanks to recombination, there are a large number of such paths, 
and genetic distance averages over these.
It is reasonable in some situations to model these lineages as random walks across the landscape;
however, the model of genetic distance that one is thus led to
is \emph{coalescence time} of the random walks, 
rather than commute time, the quantity that corresponds to resistance.
This naturally raises the question: 
Are methods that depend on effective resistances being misled?
If so, would using coalescence time do better?
To better contrast the two,
we refer to ``commute time'' rather than ``resistance'' throughout,
although the two concepts are equivalent.
\plr{CHANGE TO mostly RESISTANCE STARTING AT SIMPLE EXAMPLE}

In this paper, we contrast 
coalescence time and commute time,
and develop an inference method for movement rates on a landscape
using both as estimated from genetic distances.
As we will see,
although the two methods are conceptually similar,
commute time inference may get the wrong answer if data derive from a coalescent process,
as does real genetic data.
We also explore the question of identifiability.
Resistance-based methods are often used to predict distances based on a few layers of geographical data
(land cover, slope, etcetera).
Some methods find the combination of landscape layers to best explain genetic distance,
which allows very fine geographic resolution \citep{infer_resistance}.
Others methods such as EEMS \citep{eems} infer conductances without such prior information;
however, the geographical resolution is much coarser.
One reason for this difference in resolution
is that EEMS uses more computationally intensive Bayesian methods.
However, we show that this is a more general problem --
inference of movement rates, in either a coalescence or commute time framework,
is an ill-conditioned problem,
implying the need for regularization to obtain reliable inference.
Using maps for prior information, or restricting the geographical resolution of the results,
are two ways to do this.

@@

There are a number of ways to calculate measures of dissimilarity using genetic data \citep{distances}.
In this paper we use mean pairwise divergence:
for two genomes, it is calculated as the density of sites at which the two differ.
The genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
from the two genomes back to their most recent common ancestor at that site;
under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
pairwise divergence is therefore an unbiased estimate of the mean time to most recent common ancestor,
averaged across the genome \citep{hudson2007variance,ralphXX}.
It is natural to model this process by following the two lineages back through time
(and across space)
until they find their most recent common ancestor,
i.e., until they coalesce.
It is common to model the geographic distribution of a species
as a collection of randomly mating populations
that exchange occasional migrants,
thus discretizing geographic space.
If the genome is selectively netural,
then the motion of these lineages back through time
forms a Markov process, the \emph{structured coalescent} \citep{wakeley2005coalescent}.
In this framework, each lineage performs a random walk across the populations,
with movement probabilities depending on the flux of individuals between populations.
This provides the link between genetic distance and random walks.

@@

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$,
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.

@@

@@

>>>>

The idea of ``resistance distance'' is an important tool in the lanscape genomics toolbox.
Introduced by \citet{mcrae2006isolation},
it makes use of a mathematical equivalence between random walks and electrical resistance
\citep{resistance}:
the expected time for a continuous-time Markov chain that starts at node $x$
to first hit node $y$ and then return to $x$ (the ``commute time'')
is equal to the effective resistance between $x$ and $y$ in an electrical network
whose conductances are given by the movement rates of the Markov chain \citep{levin}.
This measure therefore averages over all possible paths through the network between the two.
For this reason, 
if assigning local conductances based on the values of some landscape variable
results in a significant positive correlation between 
genetic distances between samples from different parts of the landscape
and the effective resistances between those locations,
then this is taken as good evidence that the landscape variable is a good indicator of
where gene flow occurs \citep{mcrae2007circuit,cushman2006complex,more_examples}.
For instance,
if the conductance across a grid cell in a discretization of the landscape
is higher in flatter areas,
and this produces effective resistance values 
that are significantly correlated with genetic distances,
then one might conclude that the species in question tends to disperse more readily through flatter areas,
perhaps along river valleys.
More recent methods \citep{petkova2016visualizing,hanks2013circuit} seek to build \emph{de novo} 
a map of conductance values that produces resistances 
most strongly correlated with observed genetic distance,
and then interpret regions of low conductance in the resulting map as barriers to gene flow.

As motivated by \citet{mcrae2006isolation},
commute time (i.e., resistance distance) is a more computationally tractable approximation
to \emph{coalescence time}, a more accurate model of genetic distance data.
Genetic distances between two genomes sampled from the landscape
do derive from an average across a large number of paths between those points --
the lineages along which each segment of genetic material has been inherited
from their most recent common ancestor.
The more recent this ancestor is,
the smaller the genetic distance will be.
Thanks to recombination, there are a large number of such paths, 
and genetic distance averages over these.
It is reasonable in some situations to model these lineages as random walks across the landscape;
however, the model of genetic distance that one is thus led to
is \emph{coalescence time} of the random walks, 
rather than commute time, the quantity that corresponds to resistance.
This naturally raises the question: 
Are methods that depend on effective resistances being misled?
If so, would using coalescence time do better?

\citet{mcrae2006isolation} showed that coalescence time and resistance distance are equal
(up to shift and scaling) in isotropic landscapes such as a ring \citep{matsen_wakeley},
and found that the two were highly correlated in several test cases.
However, all these test cases involved symmetric rates of gene flow,
which is likely not the case in many situations, e.g.,
in rivers \citep{morrissey2009maintenance,sundqvist2016directional,hanks2017modeling},
source-sink dynamics \citep{source_sink},
or after population expansions \citep{expansions}.

\plr{CHANGE TO mostly RESISTANCE STARTING AT SIMPLE EXAMPLE}

In this paper, we contrast 
coalescence time and resistance distance,
and develop an inference method for movement rates on a landscape
using both as estimated from genetic distances.
As we will see,
although the two methods are conceptually similar,
commute time inference may get the wrong answer if data derive from a coalescent process,
as does real genetic data, especially in the presence of asymmetry.
We also explore the question of identifiability.
Resistance-based methods are often used to predict distances based on a few layers of geographical data
(land cover, slope, etcetera).
Some methods find the combination of landscape layers to best explain genetic distance,
which allows very fine geographic resolution \citep{infer_resistance}.
Others methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
however, the geographical resolution is much coarser.
One reason for this difference in resolution
is that EEMS uses more computationally intensive Bayesian methods.
However, we show that this is a more general problem --
inference of movement rates, in either a coalescence or commute time framework,
is an ill-conditioned problem,
implying the need for regularization to obtain reliable inference.
This explains the results of \citet{graves2013current},
who found a large area of resistance parameters
that produced equivalently good fits to genetic distance
(i.e., nonidentifiability, analogous to a flat likelihood surface).

@@

There are a number of ways to calculate measures of dissimilarity using genetic data \citep{distances}.
In this paper we use mean pairwise divergence:
for two genomes, it is calculated as the density of sites at which the two differ.
It is more common in the literature to use $F_{ST}/(1-F_{ST})$,
but the two are highly correlated for good theoretical reasons \citet{slatkin_fst,rousset1997genetic},
especially for SNP data.
The genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
from the two genomes back to their most recent common ancestor at that site;
under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
pairwise divergence is therefore an unbiased estimate of the mean time to most recent common ancestor,
averaged across the genome \citep{hudson2007variance,ralphXX}.
It is natural to model this process by following the two lineages back through time
(and across space)
until they find their most recent common ancestor,
i.e., until they coalesce.
It is common to model the geographic distribution of a species
as a collection of randomly mating populations
that exchange occasional migrants,
thus discretizing geographic space.
If the genome is selectively netural,
then the motion of these lineages back through time
forms a Markov process, the \emph{structured coalescent} \citep{wakeley2005coalescent}.
In this framework, each lineage performs a random walk across the populations,
with movement probabilities depending on the flux of individuals between populations.
This provides the link between genetic distance and random walks.

@@

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
Similar recursions are common in the literature,
going back at least to \citet{hill1972effective} (see also \citet{whitlock1997effective}).
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$,
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.

@@

\paragraph{Asymmetric dispersal}
(Review of asymmetric situations.)

\citet{hanks2017modeling} developed a method that allows asymmetry
by modeling genetic similarity as deriving from an underlying random field
whose space-time covariance is given by the covariance of the (forwards-time) population fluctuations.

@@

\paragraph{Limitations of data}
Something about microsats versus RADseq versus whole-genome.
\plr{XXX}

\paragraph{Landscapes or discrete demes?}
Other inference methods \citep{wilson2003bayesian,beerli1999maximum,beerli2010unified,hey2007integration}
work with coalescence inference, but only with a small number of loci and populations,
because of computational efficiency.
We're somewhere in between.
\plr{XXX}
